  ---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
#loading the data set and converting the date column into date class
library(lubridate)
data <- read.csv("activity.csv")
data$date <- as.Date(as.character(data$date))
str(data)
```

## What is mean total number of steps taken per day?
```{r}
library(ggplot2)
library(dplyr)
mydata <- summarise(group_by(data[c("steps","date")], date), Total.steps = sum(steps))
g <- ggplot(mydata, aes(Total.steps))
p <- g + geom_histogram(col = "blue", fill = "green") +  xlab("Total number of steps") + ylab("Frequencies per day") + ggtitle("Total number of steps taken per day")
p
```
  
## Mean and Median of total number of steps taken per dat
  
```{r}
#Mean of total steps taken each day
print(paste("Mean of the provided dataset", mean(mydata$Total.steps , na.rm = T)))
#Median of total steps taken each day
print(paste("Median of the provided dataset", median(mydata$Total.steps , na.rm = T)))
```

## What is the average daily activity pattern?
```{r}
mydat2 <- summarise(group_by(na.omit(data[c("steps","interval")]) , interval), AvgSteps = mean(steps))
plot(mydat2$interval, mydat2$AvgSteps, type = "l", xlab = "Interval(min)", ylab = "AverageSteps", main = "Average daily activity pattern", col = "red")

```
  
  
## Find the interval with the maximum amount of average activity:
```{r}
Intervalmax <- mydat2$interval[which.max(mydat2$AvgSteps)]
print(paste('Interval with largest mean number of steps per day is = ', Intervalmax))
```

## Imputing missing values
```{r}
#No. of NA values in the original data set is
print(paste('No. of rows with NAs value are = ',nrow(data[is.na(data$steps),])))
#Imputing the NAs with the average of daily activity
missingInd <- which(is.na(data$steps))
newdata <- data
for (index in missingInd) {
  newdata$steps[index] <- ceiling(mydat2$AvgSteps[mydat2$interval == newdata$interval[index]])
}
```
  
    
## Creating a histogram of the new dataset
```{r}
newmydata <- summarise(group_by(newdata[c("steps","date")], date), Total.steps = sum(steps))
newg <- ggplot(newmydata, aes(Total.steps))
newp <- newg + geom_histogram(col = "purple", fill = "orange") +  xlab("Total number of steps") + ylab("Frequencies per day") + ggtitle("Total number of steps taken per day")
newp
```
  
    
## Calculating the mean and median of the total number of steps taken per day of the imputed dataset:
```{r}
#Mean of total steps taken each day
print(paste("Mean of the imputed dataset",mean(newmydata$Total.steps)))
#Median of total steps taken each day
print(paste("Median of the imputed dataset",median(newmydata$Total.steps)))
```
The median and mean of the total steps per day increased due to the imputed values for NAs.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
slavedays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
newdata$Type.day <- as.factor(ifelse(weekdays(newdata$date) %in% slavedays, "weekday", "weekend"))
fg <- ggplot(newdata, aes(y = steps, x = interval))
fp <- fg + geom_line() + facet_grid(rows = vars(Type.day)) + labs(x = "Time interval", y = "Average Steps", title = "Differences in activity patterns between weekdays and weekends")
fp
```